Fast time-domain electromagnetic modelling in the frequency domain

### 2020-06-30 ManuscriptTime.tex

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. We present a
fast frequency-to-time domain method using either the digital linear filter
method or the logarithmic fast Fourier transform together with careful
selection of frequencies and interpolation. 15 to 20 frequencies are generally
enough with this methodology to cover a wide range of offsets in
one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three points results in much faster computation
times with speedup-factors of ten and more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and their corresponding
grids are carefully chosen.
%
\end{abstract}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not just in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, or geothermal sources
\citep{WGC.15.Girard}. While in the deep marine environment current sources
with few frequencies are used, transient measurements are more common in the
shallow marine environment and on land. One of the main reasons is the
dominance of the airwave in shallow marine and terrestrial measurements, which
can be better separated in the time domain.

CSEM is usually divided into frequency- and time-domain methods, depending
whether the source signal is a continuous waveform, such as a sine, or a finite
waveform, such as a pseudo-random binary sequence (PRBS). A numerical
comparison of the two methods is given in \cite{GP.13.Conell}. Acquired CSEM
data are subsequently often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time (see,
e.g., \cite{SG.10.Borner, SG.05.Avdeev} for comprehensive overviews of CSEM
codes). However, under certain conditions it can be competitive to model
time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. The conditions are: a sufficiently powerful solver,
appropriate frequency selection and interpolation, and an automated gridding,
for which they used the multi-frequency CSEM approach presented by
\cite{GEO.07.Plessix}. We build upon these results but improve the run time
from hours to minutes. The main reasons for this significant speed-up are an
adaptive, frequency-dependent gridding scheme that minimizes the required cells
in each dimension, and a logarithmic Fourier transform such as digital linear
filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast Fourier transform
\citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency to the time
domain. The latter makes it also possible to only use the imaginary part of the
frequency-domain response, which has advantages when it comes to interpolation.

In the next section, we briefly review the methods and highlight their
advantages and shortcomings. This is followed by an outline of our changes to
the Fourier transform and the adaptive gridding. Finally, we demonstrate the
efficiency of the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains, as it is done, for instance, in \simpeg \citep{CAG.15.Cockett}.
Another method is to have Maxwell's equation only implemented in one domain,
and use Fourier transforms to go to the other domain. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, which translates into a
significant faster computation times. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be $2-3$ times bigger than $\Delta_\mathrm{min}$.
The smallest cells have to be around the source; in the marine case the minimum
skin depth is therefore defined by the conductivity of seawater. However, this
yields quite large cells for low frequencies, so special care has to be taken
around the source. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the boundary zone outside of the
area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it looses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes. It is important to note that while we show the transform from frequency
to time domain, the conclusions regarding the Fourier transform can equally
well be applied to the transform from time to frequency domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \cite{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, personal
communication).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time convention is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \emg3d has implemented known improvements such as line-relaxation and
semicoarsening \citep{ECCFD.06.Jonsthovel}, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
18.04 and Python 3.7. For the semi-analytical computations of layered models we
use the open-source (Apache License 2.0) code \empymod
\citep{GEO.17.Werthmuller}.

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the faster we obtain the
time-domain result. We decided to use a regular spacing of frequencies on a
log-scale, rather than an adaptive scheme. This approach is favourable if a
wide range of offsets is needed, as the required frequencies change with offset
and an adaptive frequency selection is therefore often tailored to a single
offset. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale, and they only require either the real or the
imaginary part of the frequency-domain response. We can take advantage of that
by using only the imaginary part of the frequency-domain response. The
imaginary part has the characteristics that its value goes to zero when the
frequency goes to zero or to infinity, with the advantage that knowing the
endpoints makes it possible to convert the extrapolation of missing frequencies
into interpolation.

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\plot{GUI-FFTLog-lin}{width=\tcwidth}{
  Interactive frequency selection for a user-provided layered model (the shown
  model parameters are described in the text). This example shows the impulse
  response at an offset of 5\,km, for which it uses FFTlog from 0.001\,Hz to
  10\,Hz with five frequencies per decade. Note that the error is clipped for
  values smaller than 0.01\,\% and bigger than 100\,\%.
}
%
The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the precise results
obtained from \empymod. The blue circles indicate the actually computed
responses and the black dots the frequencies which are interpolated or set to
zero. The resulting time-domain response has a relative error of less than
1\,\% everywhere except for very early times.

@@

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the less frequencies are needed. Another important
factor is how to interpolate and extrapolate from the computed frequencies to
the frequencies required for the Fourier transform. For the FFTLog only
extrapolation for higher and lower frequencies is required. The EM response
becomes highly oscillatory for high frequencies, which makes it very hard to
extrapolate the response to frequencies $f>f_\mathrm{max}$. However, if
$f_\mathrm{max}$ is chosen appropriately, the importance of higher frequencies
for the Fourier transform can be neglected and we can set those responses to
zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be changed to an
interpolation by assuming a zero imaginary response at a very low frequency,
and we then use PCHIP to interpolate the missing frequencies. In the case of
the DLF method we also have to interpolate in between the computed frequencies,
for which we found it better to use a cubic spline. As can be seen from Figures
\ref{fig:GUI-FFTLog-lin} and \ref{fig:GUI-DLF-lin}, using the FFTLog or the DLF
with the same actually computed frequencies results in very similar responses.

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\plot*{fullspace}{width=\textwidth}{
  (a) Frequency- and (b) time-domain results for the homogeneous space model.
  The red lines are the analytical solutions, the blue circles are the actually
  computed responses with \emg3d, the black dots are the interpolated
  responses, and the dashed black line the obtained time-domain response. The
  errors are clipped for values smaller than 0.1\,\% and bigger than 100\,\%.
}
%
Most of the computed values stay below a relative error of 1\,\%, our
chosen adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result from \emg3d, on top of the
red line which is the analytical result. The relative error is mostly below
1\,\%, except for early times. However, for practical reasons that is more than
enough. Figure~\ref{fig:fullspace-log} shows the same on a logarithmic scale,
with times up to 10 seconds. It clearly shows that if later times are required,
we would need to adjust our Fourier transform parameters.
%
\plot{fullspace-log}{width=\cwidth}{
  Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic scale. To
  improve later times we would have to compute lower frequencies; to improve
  earlier times we would have to compute more frequencies per decade to get a
  better resolution.
}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; $3-4$ is often
enough.

@@

This massive speed-up has a couple of reasons. Computers have become more
powerful in the last twelve years, and the codes were run on different
computers. A quick test with the old scripts on our test machine shows that it
would roughly run 2-3 times faster, therefore somewhere between 1 and 2 hours.
The more important facts besides different hardware are: (1) we only used 14
frequencies instead of the 26 frequencies between 0.01 and 100\,Hz of the
original; (2) our adaptive gridding used significantly less cells
($f$-dependent) in comparison to the fixed \num{2097152} cells ($128^3$) used
in the original example. We did not see a significant difference in the speed
of the actual codes, where the kernel-algorithm of the two implementations is
the same, but in the original example it is implemented in Matlab/C, whereas
\emg3d is written in Python/Numba (Numba is a just-in-time compiler for Python
code, \citealp{Numba}).

@@

This research was conducted within the Gitaro.JIM project funded through
MarTERA as part of Horizon 2020, a funding scheme of the European Research Area
(ERA-NET Cofund, \href{https://www.martera.eu}{martera.eu}).

@@

\emph{Note to editors and reviewers: The data is NOT YET on Zenodo, we will
  upload there the final versions and link the right DOI for the final
  manuscript. Until then everything can be found on GitHub at
  \href{https://github.com/empymod/article-TDEM}%
  {github.com/empymod/article-TDEM}.}

>>>>

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. The fast
frequency-to-time domain method uses either the digital linear filter method
or the logarithmic fast Fourier transform together with a careful selection of
frequencies and interpolation. This methodology requires typically 15 to 20
frequencies to cover a wide range of offsets in one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three points results in much smaller computation
times with speedup factors of ten or more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and the corresponding grids
are carefully chosen.
%
\end{abstract}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or geological purposes \citep{NAT.19.Johanson}. While
current sources with a few frequencies are used in the deep marine environment,
transient measurements are more common in the shallow marine environment and on
land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis, SEG.07.Avdeeva}. One
of the main reasons is the dominance of the airwave in shallow marine and
terrestrial measurements, which can be better separated in the time domain.
CSEM is usually divided into frequency- and time-domain methods, depending
whether the source signal is a continuous waveform, such as a sine, or a finite
waveform, such as a pseudo-random binary sequence (PRBS). A numerical
comparison of the two methods is given by \cite{GP.13.Conell}. Acquired CSEM
data are subsequently often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's, with a
good state of the art provided by \cite{B.SEG.99.Oristaglio}. A second wave
occurred during the CSEM boom in the hydrocarbon industry in the early 21th
century, see, e.g., \cite{SG.05.Avdeev} and \cite{SG.10.Borner} for
comprehensive overviews. Recent publications span the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences \citep{CAG.13.Sommer}, finite
elements \citep{GJI.13.Grayver}, and finite volumes \citep{GEO.14.Jahandari}.
Besides the different methods there are also many different types of
discretisation. Commonly used, besides rectilinear Yee grids
\citep{IEEE.66.Yee} as used by our code, are Lebedev grids
\citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedral \citep{CAG.17.Cai}. The computational requirement to model CSEM
data also depends heavily on the technique which is used to solve the system,
where there are direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be competitive to model time-domain data with a
frequency-domain code, as shown by \cite{GEO.08.Mulder}. The conditions are: a
sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

In the next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains, as it is done, for instance, in \simpeg \citep{CAG.15.Cockett}.
Another approach is to have Maxwell's equation only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the time
domain to the frequency domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies, translating into a significant
reduction in computation time. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the boundary zone outside of the
area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes. It is important to note that while we show the transform from frequency
to time domain, the conclusions regarding the Fourier transform can equally
well be applied to the transform from time to frequency domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \citep{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, private
communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \emg3d has implemented known improvements such as line-relaxation and
semicoarsening \citep{ECCFD.06.Jonsthovel}, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
18.04 and Python 3.7. For the semi-analytical computations of layered models we
use the open-source (Apache License 2.0) code \empymod
\citep{GEO.17.Werthmuller}.

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. We decided to use a regular spacing of frequencies on a
log-scale, rather than an adaptive scheme. This approach is favourable if a
wide range of offsets is needed, as the required frequencies change with offset
and an adaptive frequency selection is therefore often tailored to a single
offset. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale, and they only require either the real or the
imaginary part of the frequency-domain response. We can take advantage of that
by using only the imaginary part of the frequency-domain response. The
imaginary part goes to zero when the frequency goes to zero or to infinity,
with the advantage that knowing the endpoints makes it possible to convert the
extrapolation of missing frequencies into interpolation.

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\plot{GUI-FFTLog-lin}{width=\tcwidth}{
  Interactive frequency selection for a user-provided layered model (the shown
  model parameters are described in the text). This example shows the impulse
  response at an offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to
  10\,Hz with five frequencies per decade. Note that the error is clipped for
  values smaller than 0.01\,\% and bigger than 100\,\%.
}
%
The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m at 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the precise results
obtained from \empymod. The blue circles indicate the actually computed
responses and the black dots the frequencies which are interpolated or set to
zero. The resulting time-domain response has a relative error of less than
1\,\% everywhere except for very early times.

@@

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the less frequencies are needed. Another important
factor is how to interpolate and extrapolate from the computed frequencies to
the frequencies required for the Fourier transform. For the FFTLog only
extrapolation for higher and lower frequencies is required. The EM response
becomes highly oscillatory for high frequencies, which makes it very hard to
extrapolate the response to frequencies $f>f_\mathrm{max}$. However, if
$f_\mathrm{max}$ is chosen judiciously, the importance of higher frequencies
for the Fourier transform can be neglected and we can set those responses to
zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be changed to an
interpolation by assuming a zero imaginary response at zero frequency, and we
then use PCHIP to interpolate the missing frequencies (as we work on a
logarithmic scale we cannot choose 0\,Hz, but instead take $10^{-100}$\,Hz). In
the case of the DLF method we also have to interpolate in between the computed
frequencies, for which we found it better to use a cubic spline. As can be seen
from Figures \ref{fig:GUI-FFTLog-lin} and \ref{fig:GUI-DLF-lin}, using the
FFTLog or the DLF with the same actually computed frequencies results in very
similar responses.

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\plot*{fullspace}{width=\textwidth}{
  (a) Frequency- and (b) time-domain results for the homogeneous space model.
  The red lines are the analytical solutions, the blue circles are the actually
  computed responses with \emg3d, the black dots are the interpolated
  responses, and the dashed black line the obtained time-domain response. The
  errors are clipped for values smaller than 0.1\,\% and bigger than 100\,\%.
}
%
Most of the computed values stay below a relative error of 1\,\%, our
chosen adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result from \emg3d, on top of the
red line which is the analytical result. The relative error is mostly below
1\,\%, except for early times. However, for practical reasons that is more than
enough. Figure~\ref{fig:fullspace-log} shows the same on a logarithmic scale,
with times up to 10 seconds. It clearly shows that if later times are required,
we would need to adjust our Fourier transform parameters.
%
\plot{fullspace-log}{width=\cwidth}{
  Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic scale. To
  improve later times we would have to compute lower frequencies; to improve
  earlier times we would have to compute more frequencies per decade to get a
  better resolution.
}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.

@@

This massive speed-up has a couple of reasons. Computers have become more
powerful in the last twelve years, and the codes were run on different
computers. A quick test with the old scripts on our test machine shows that it
would roughly run 2--3 times faster, therefore somewhere between 1 and 2 hours.
The more important facts besides different hardware are: (1) we only used 14
frequencies instead of the 26 frequencies between 0.01 and 100\,Hz of the
original; (2) our adaptive gridding used significantly less cells
($f$-dependent) in comparison to the fixed \num{2097152} cells ($128^3$) used
in the original example. We did not see a significant difference in the speed
of the actual codes, where the kernel-algorithm of the two implementations is
the same, but in the original example it is implemented in Matlab/C, whereas
\emg3d is written in Python/Numba (Numba is a just-in-time compiler for Python
code, \citealp{Numba}).

@@

This research was conducted within the Gitaro.JIM project funded through
MarTERA as part of Horizon 2020, a funding scheme of the European Research Area
(ERA-NET Cofund, \href{https://www.martera.eu}{martera.eu}).
An early idea of this manuscript was presented at \cite{EAGE.20.Werthmuller}.

@@

\emph{Note to editors and reviewers: The data are NOT YET on Zenodo, we will
  upload there the final versions and link the right DOI for the final
  manuscript. Until then everything can be found on GitHub at
  \href{https://github.com/empymod/article-TDEM}%
  {github.com/empymod/article-TDEM}.}
